Response to decision letter

### 2023-05-12 manuscript/manuscript.Rmd
add EJM edits on discussion

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms  [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Alternative source chronograms available for the same taxa have potentially been estimated implementing calibrations very differently.
In the Fringillidae example above, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->

Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates [@sauquet2013practical].
In addition, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], this inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.

@@

While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife also summarizes chronograms using secondary calibrations (ages coming from other chronograms).
The scientific community has generally more confidence in chronograms generated from a single analysis, where fossil data constitute the primary source of calibrations [@sauquet2013practical; @schenk2016sec; @antonelli2017supersmart].
@graur2004reading cautioned on the increased error and uncertainty in estimated node ages when using secondary calibrations in dating analyses, and more recently, @schenk2016sec found that divergence times obtained using secondary calibrations with Bayesian inference methods are significantly younger than those obtained using primary calibrations. While this trend still needs to be evaluated with alternative methods of divergence time estimation,
 many efforts and discussions on best practices for implementing primary calibrations have been advanced in recent years,
for example, using fossils that have been widely discussed and curated as calibrations to date other trees and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->

Obtaining vetted primary calibrations from the litterature is another bottleneck for facilitating obtention of chronograms.
The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications [@ksepka2015fossil]. This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis. <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
Despite the wealth of information displayed in the database, users still need the expertise to locate the equivalent node in their phylogenies to be dated.
Open databases for reuse of fossil calibration data still have room to grow and should be a great addition to the phylogenetics tool set.

<!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}-->

It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases [@sauquet2013practical].
Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses. It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20). The only exception to this was observed on cross validation results of the @burns2014phylogenetics chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs. S1, S5).

@@

Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data. These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].

>>>>

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence [@sauquet2013practical].
Finally, placement of calibrations also affects uncertainty of node age estimates. For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.

DateLife's summary chronograms are intended to represent all variation in estimated node ages from source chronograms. Node age distribution ranges allow to visually explore ages from source chronograms individually and contextualize and compare them against other chronograms. Researchers that wish to use summary chronograms in downstream evolutionary analysis may select multiple trees sampled from the summary distribution of node ages, to account for variation in source chronograms.

@@

While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis [@shaul2002playing; @graur2004reading; @sauquet2012testing; @sauquet2013practical; @garzon2015incompatible; @schenk2016sec; @antonelli2017supersmart].
However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply  calibrations appropriately [@ksepka2011synthesizing; @hipsley2014beyond]. One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees [@ksepka2011synthesizing; @sauquet2013practical], and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications [@ksepka2015fossil]. This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.
 <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->

In some cases, secondary calibrations might represent the only information available (e.g., groups with a poor fossil record). Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations. It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before [@shaul2002playing; @graur2004reading; @sauquet2013practical]. <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations. In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figures S2-S20). We also note that nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies. The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].

We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying]. It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s]; with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic]; and with methods that summarize age data unto a tree topology, such as DateLife.

@@

Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data. These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions [@cusimano2012new; @chang2020estimating], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].

### 2023-05-12 notes-effects-downstream-analysis-review.Rmd
add EJM edits on discussion

Adding branch lengths sampled from a birth-death model in the absence of genetic data
improves insight from phylogeny-based analyses of diversification rates [@chang2020estimating].
@chang2020estimating state that stochastic politomy resolution alleviate low statistical power to
identify shifts in diversification rate with methods that account for sampling fractions.
@chang2020estimating "demonstrate theoretical concerns with the widespread use of analytical
sampling corrections for sparsely sampled phylogenies of higher taxonomic groups. In particular,
corrections based on “sampling fractions” can lead to low statistical power to infer rate variation
when it is present, depending on the likelihood function used for inference. In the extreme, the sampling
fraction correction can lead to spurious patterns of diversification that are driven solely by unbalanced
sampling across the tree in concert with low overall power to infer shifts."

@@

@sun2020estimating used a 50% species level sampled phylogeny of the family Cucurbitaceae to estimate
speciation rates with two parametric methods (RPANDA and BAMM) and one semi-parametric method (DR) and
tested the effect of using incompletely sampled phylogenies on one hand, and incomplete phylogenies that
have been made near complete, using a stochastic politomy resolver based on taxonomy, on the other hand
(with two different methods: 1) adding missing species at random following a backbone taxonomy, and 2)
adding missing species to a genus level taxonomy based on a taxonomy resolver only). When using incomplete
phylogenies, parametric methods tend to overestimate speciation rate as sampling gets worse (less complete),
while semi-parametric methods tend to underestimate speciation rate as sampling gets worse. When using
taxonomically resolved complete phylogenies, only one parametric method overestimates speciation rate the
more incomplete original sampling was. RPANDA, BAMM tip rates and DR all underestimate speciation rates as
original sampling was more incomplete.
"Adding backbone taxa tends to result in the inference of spurious early bursts of diversification", bias in
sampling ancestral nodes, failure to sample recent divergences or pushing them back in relative time (making
recent divergences older than expected)
"failure to detect recent bursts of speciation and instead inferring higher rates of evolution at earlier time
intervals"

>>>>

Adding branch lengths sampled from a birth-death model in the absence of genetic data
improves insight from phylogeny-based analyses of diversification rates [@chang2020estimating].
@chang2020estimating state that stochastic polytomy resolution alleviate low statistical power to
identify shifts in diversification rate with methods that account for sampling fractions.
@chang2020estimating "demonstrate theoretical concerns with the widespread use of analytical
sampling corrections for sparsely sampled phylogenies of higher taxonomic groups. In particular,
corrections based on “sampling fractions” can lead to low statistical power to infer rate variation
when it is present, depending on the likelihood function used for inference. In the extreme, the sampling fraction correction can lead to spurious patterns of diversification that are driven solely by unbalanced sampling across the tree in concert with low overall power to infer shifts."

@@

@sun2020estimating used a 50% species level sampled phylogeny of the family Cucurbitaceae to estimate
speciation rates with two parametric methods (RPANDA and BAMM) and one semi-parametric method (DR) and
tested the effect of using incompletely sampled phylogenies on one hand, and incomplete phylogenies that
have been made near complete, using a stochastic polytomy resolver based on taxonomy, on the other hand
(with two different methods: 1) adding missing species at random following a backbone taxonomy, and 2)
adding missing species to a genus level taxonomy based on a taxonomy resolver only). When using incomplete
phylogenies, parametric methods tend to overestimate speciation rate as sampling gets worse (less complete),
while semi-parametric methods tend to underestimate speciation rate as sampling gets worse. When using
taxonomically resolved complete phylogenies, only one parametric method overestimates speciation rate the
more incomplete original sampling was. RPANDA, BAMM tip rates and DR all underestimate speciation rates as
original sampling was more incomplete.
"Adding backbone taxa tends to result in the inference of spurious early bursts of diversification", bias in
sampling ancestral nodes, failure to sample recent divergences or pushing them back in relative time (making
recent divergences older than expected)
"failure to detect recent bursts of speciation and instead inferring higher rates of evolution at earlier time
intervals"
